---
title: "scRNAseq analysis of Levesque’s data"
author: "Leonor Schubert, Jonathan Haab, Flavio Rump"
date: "22 12 2020"
output: html_document
---
from pipeline example described by Pierre-Luc Germain, course sta426 UZH

# Loading necessary libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scran)
  library(scater)
  library(batchelor)
  library(scDblFinder)
  library(sctransform)
  library(muscat)
  library(SEtools)
  library(cowplot)
  library(BiocParallel)
  library(ComplexHeatmap)
  
  library(DropletUtils) # for read10xCounts
  library(Matrix) # to handle sparse matrix
  library(BiocNeighbors) # for kNN graph
  library(igraph) # for cluster_louvain
})
```

```{r sourceCode,echo=TRUE}
# set local path
local.path <- getwd()
setwd(local.path)

# use source() function to source the functions we want to execute
source("functions/functions_qc.R")
source("functions/functions_analysis.R")
```

# Preprocessing & Clustering

## Loading the data

```{r data-loading}
## data can be downloaded from : http://imlspenticton.uzh.ch/dump/files_for_levesque.tar

patient1_HS.path <- file.path("data", "patient1_HS")

# read in filtered data
fnameHS <- file.path(patient1_HS.path, "outs/filtered_feature_bc_matrix")

#define data we want to look at
sample_names <- list("patient1_HS", "patient1_SCC", "patient2_HS", "patient2_AK")
patient1_HS.path <- file.path("data", "patient1_HS")
patient1_SCC.path <- file.path("data", "patient1_SCC")
patient2_HS.path <- file.path("data", "patient2_HS")
patient2_AK.path <- file.path("data", "patient2_AK")


#read 10X data
# read in filtered data and create list of SingleCellExperiment objects
paths <- list(patient1_HS.path, patient1_SCC.path, patient2_HS.path, patient2_AK.path)
#sces <- lapply(paths, function(i) read10xCounts(file.path(i, "outs/filtered_feature_bc_matrix"), col.names = TRUE))

#read our own data
sces <- list()
#sces[[1]] <- readRDS('filtered_clean_data/patient1_HS_cleanData_sce.RDS')
#sces[[2]] <- readRDS('clean_data/patient1_SCC_cleanData_sce.RDS')
#sces[[3]] <- readRDS('clean_data/patient2_HS_cleanData_sce.RDS')
#sces[[4]] <- readRDS('clean_data/patient2_AK_cleanData_sce.RDS')

sces[[1]] <- processCellRangerOutput("patient1_HS", 100000, TRUE, TRUE, TRUE, 6)
sces[[2]] <- processCellRangerOutput("patient1_SCC", 100000, TRUE, TRUE, TRUE, 6)
sces[[3]] <- processCellRangerOutput("patient2_HS", 100000, TRUE, TRUE, TRUE, 6)
sces[[4]] <- processCellRangerOutput("patient2_AK", 100000, TRUE, TRUE, TRUE, 6)

```

```{r split-data}

sces <- lapply(sces, split_data)

# only select the gene expression data, drop the antibody capture
sce.patient1_HS <- sces[[1]]
sce.patient1_SCC <- sces[[2]]
sce.patient2_HS <- sces[[3]]
sce.patient2_AK <- sces[[4]]

# add name to sce for output graph
attr(sce.patient1_HS, "name") <- "Patient1 HS"
attr(sce.patient1_SCC, "name") <- "Patient1 SCC"
attr(sce.patient2_HS, "name") <- "Patient2 HS"
attr(sce.patient2_AK, "name") <- "Patient2 AK"
```

## Normalization & reduction

```{r vst, include=FALSE}

results <- variance_stabilization(sce.patient1_HS, 2000)
sce.patient1_HS <- results[[1]]
hvg.patient1_HS <- results[[2]]

results <- variance_stabilization(sce.patient1_SCC, 2000)
sce.patient1_SCC <- results[[1]]
hvg.patient1_SCC <- results[[2]]

results <- variance_stabilization(sce.patient2_HS, 2000)
sce.patient2_HS <- results[[1]]
hvg.patient2_HS <- results[[2]]

results <- variance_stabilization(sce.patient2_AK, 2000)
sce.patient2_AK <- results[[1]]
hvg.patient2_AK <- results[[2]]
```

ADD A NICE TABLE WITH ALL THE MARKER USED (and their source) AND EXPLAINED WHICH WEREN'T FOUND
also explain that we decided to add the known markers from the litterature into the group of genes used to produce a low-dimensional representation of the data even when they were not part of the most variable genes

```{r pca}
# run PCA

# using known genes
# genes not found in the data : "DCDN" "WISP2" "SEPP1" "TYP1" 
# genes linked to pericytevSMC : "ACTA2","TAGLN"
# were removed from the following list
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","LUM","MMP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","CDH19","NGFR","UBE2C","PCNA")

sce.patient1_HS <- sc_PCA(sce.patient1_HS, hvg.patient1_HS, known_markers)
sce.patient1_SCC <- sc_PCA(sce.patient1_SCC, hvg.patient1_SCC, known_markers)
sce.patient2_HS <- sc_PCA(sce.patient2_HS, hvg.patient2_HS, known_markers)
sce.patient2_AK <- sc_PCA(sce.patient2_AK, hvg.patient2_AK, known_markers)

```

## Clustering

```{r clustering}

results <- sc_cluster(sce.patient1_HS)
sce.patient1_HS <- results[[1]]
g.patient1_HS <- results[[2]]

results <- sc_cluster(sce.patient1_SCC)
sce.patient1_SCC <- results[[1]]
g.patient1_SCC <- results[[2]]

results <- sc_cluster(sce.patient2_HS)
sce.patient2_HS <- results[[1]]
g.patient2_HS <- results[[2]]

results <- sc_cluster(sce.patient2_AK)
sce.patient2_AK <- results[[1]]
g.patient2_AK <- results[[2]]

```

# Cluster annotation

## Known markers

```{r}
# genes not found in our data : "DCDN" "WISP2" "SEPP1" "TYP1" 
# were removed from the list below
genes <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
  # KERATINOCYTE
  keratinocyte = c("DSC3","DSP","LGALS7"),
  #keratinocyte_basal = c("KRT5","KRT14","COL17A1"),
  #keratinocyte_suprabasal = c("KRT1","KRT10"),
  #keratinocyte_differentiated = c("LOR", "SPINK5"),
  #keratinocyte_ORS = c("KRT6B"),
  #keratinocyte_channel = c("GJB2","GJB6","ATP1B1"),
  #keratinocyte_sebaceous_gland = c("MGST1","FASN"),
  #keratinocyte_sweat_gland  = c("DCD","AQP5"),
  # FIBROBLAST
  fibroblast  = c("LUM","MMP2"),
  # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
  #fibroblast_secretory_reticular = c(SLPI","CTHRC1","MFAP5","TSPAN8"),
  #fibroblast_proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
  #fibroblast_secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
  #fibroblast_mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1"),
  # PERICYTE & vSMC
  #pericytevSMC  = c("ACTA2","TAGLN"), #vSMC : vascular smooth muscel cell
  #pericytevSMC_pericyte  = c("RGS5"),
  #pericytevSMC_vSMC  = c("MYL9","TPM2","RERGL"),
  # ENDOTHELIAL CELL
  endothelial  = c("PECAM1","SELE","CLDN5","VWF"),
  #endothelial_lymphatic  = c("PROX1","LYVE1"),
  # MYELOID CELL
  myeloid  = c("HLA-DRA","FCER1G","TYROBP","AIF1"),
  #myeloid_dendritic  = c("CD1C"),
  #myeloid_langerhans  = c("CD207"),
  #myeloid_macrophage = c("CD68","RNASE1","ITGAX"),
  # LYMPHOCYTE
  lymphocyte  = c("CD3D","CD3E","CD52","IL7R"),
  # MELANOCYTE
  melanocyte  = c("DCT","MLANA","PMEL"),
  # SCHWANN CELL
  schwann  = c("CDH19","NGFR"),
  # MITOTIC CELL
  mitotic  = c("UBE2C","PCNA")
)


rownames(sce.patient1_HS) <- convert_rownames(sce.patient1_HS, g.patient1_HS, genes)
results <- annotate_cells(sce.patient1_HS, g.patient1_HS, genes)
sce.patient1_HS <- results[[1]]
km.patient1_HS <- results[[2]]

rownames(sce.patient1_SCC) <- convert_rownames(sce.patient1_SCC, g.patient1_SCC, genes)
results <- annotate_cells(sce.patient1_SCC, g.patient1_SCC, genes)
sce.patient1_SCC <- results[[1]]
km.patient1_SCC <- results[[2]]

rownames(sce.patient2_HS) <- convert_rownames(sce.patient2_HS, g.patient2_HS, genes)
results <- annotate_cells(sce.patient2_HS, g.patient2_HS, genes)
sce.patient2_HS <- results[[1]]
km.patient2_HS <- results[[2]]

rownames(sce.patient2_AK) <- convert_rownames(sce.patient2_AK, g.patient2_AK, genes)
results <- annotate_cells(sce.patient2_AK, g.patient2_AK, genes)
sce.patient2_AK <- results[[1]]
km.patient2_AK <- results[[2]]

```

## Pseudo-bulk aggregation

```{r}
#Patient 1 HS
#km.patient1_HS
#km.patient1_SCC
#rowData(sce.patient1_HS[c("keratinocyte", "fibroblast", "endothelial", "myeloid", )])$Symbol

results <- pseudobulk(sce.patient1_HS, km.patient1_HS)

sce.patient1_HS <- results[[1]]
pb.patient1_HS <- results[[2]]
mat.patient1_HS <- results[[3]]
cl2.patient1_HS <- results[[4]]

#Patient 1 SCC
results <- pseudobulk(sce.patient1_SCC, km.patient1_SCC)

sce.patient1_SCC <- results[[1]]
pb.patient1_SCC <- results[[2]]
mat.patient1_SCC <- results[[3]]
cl2.patient1_SCC <- results[[4]]

#Patient 2 HS
results <- pseudobulk(sce.patient2_HS, km.patient2_HS)

sce.patient2_HS <- results[[1]]
pb.patient2_HS <- results[[2]]
mat.patient2_HS <- results[[3]]
cl2.patient2_HS <- results[[4]]

#Patient 2 AK
results <- pseudobulk(sce.patient2_AK, km.patient2_AK)

sce.patient2_AK <- results[[1]]
pb.patient2_AK <- results[[2]]
mat.patient2_AK <- results[[3]]
cl2.patient2_AK <- results[[4]]

```

# Sub population
```{r keratinocyte}

known_markers.kera <- c("KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5")

#annotate the subclusters
  genes.kera <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
    # KERATINOCYTE
    #keratinocyte = c("DSC3","DSP","LGALS7"),
    basal = c("KRT5","KRT14","COL17A1"),
    suprabasal = c("KRT1","KRT10"),
    differentiated = c("LOR", "SPINK5"),
    ORS = c("KRT6B"),
    channel = c("GJB2","GJB6","ATP1B1"),
    sebaceous_gland = c("MGST1","FASN"),
    sweat_gland  = c("DCD","AQP5")
    )
  
  
# TODO : find a way to avoid giving both known_merkers.kera and genes.kera since both contain the 
  # same information in essence. Maybe will be fixed by the first todo
sce.patient1_HS.kera <- cluster_subtype(sce.patient1_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient1_SCC.kera <- cluster_subtype(sce.patient1_SCC, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_HS.kera <- cluster_subtype(sce.patient2_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_AK.kera <- cluster_subtype(sce.patient2_AK, "keratinocyte", known_markers.kera, genes.kera)

```

```{r kera_dyn}
genes.dyn <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0092867420306723
  keratinocyte_basal = "COL17A1",
  keratinocyte_cycling = "MKI67",
  keratinocyte_differentiating = "KRT1"
  )

dynamic_plot(sce.patient1_HS.kera, g.patient1_HS.kera, genes.dyn)
dynamic_plot(sce.patient1_SCC.kera, g.patient1_SCC.kera, genes.dyn)
dynamic_plot(sce.patient2_HS.kera, g.patient2_HS.kera, genes.dyn)
dynamic_plot(sce.patient2_AK.kera, g.patient2_AK.kera, genes.dyn)

```

```{r fibroblast}

known_markers.fibro <- c("SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1")

genes.fibro <- list(
   #fibroblast  = c("LUM","MMP2"),
#   # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
   fibroblast_secretory_reticular = c("SLPI","CTHRC1","MFAP5","TSPAN8"),
   fibroblast_proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
   fibroblast_secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
   fibroblast_mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1")
)

sce.patient1_HS.fibro <- cluster_subtype(sce.patient1_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient1_SCC.fibro <- cluster_subtype(sce.patient1_SCC, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_HS.fibro <- cluster_subtype(sce.patient2_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_AK.fibro <- cluster_subtype(sce.patient2_AK, "fibroblast", known_markers.fibro, genes.fibro)

```




#######################################################################################
```{r check_markers, eval=FALSE}
# The following code was used to check which genes were missing from in the data
# the known_markers are from the litterature

data_markers <- rowData(sce.patient1_HS)$Symbol
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","DCDN","LUM","MMP2","WISP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","ACTA2","TAGLN","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","SEPP1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","TYP1","CDH19","NGFR","UBE2C","PCNA")


idx_notfound <- which(!(known_markers %in% rowData(sce.patient1_HS)$Symbol))

print(paste("Done:", length(known_markers)-length(idx_notfound), "/", length(known_markers), "markers found"))
cat("Genes missing in dataset: ", known_markers[idx_notfound])
```
