---
title: "scRNAseq analysis of Levesque’s data"
author: "Leonor Schubert, Jonathan Haab, Flavio Rump"
date: "22 12 2020"
bibliography: references.bib
output: html_document
---
from pipeline example described by Pierre-Luc Germain, course sta426 UZH

# Loading necessary libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scran)
  library(scater)
  library(batchelor)
  library(scDblFinder)
  library(sctransform)
  library(muscat)
  library(SEtools)
  library(cowplot)
  library(BiocParallel)
  library(ComplexHeatmap)
  
  library(DropletUtils) # for read10xCounts
  library(Matrix) # to handle sparse matrix
  library(BiocNeighbors) # for kNN graph
  library(igraph) # for cluster_louvain
  
  library(devtools)
  library(easyGgplot2)
  
})
```

```{r sourceCode,echo=TRUE}
# set local path
local.path <- getwd()
setwd(local.path)

# use source() function to source the functions we want to execute
source("functions/functions_qc.R")
source("functions/functions_analysis.R")
```

# Preprocessing & Clustering

## Loading the data

We look at both the raw and unfiltered data from the cell ranger output, to see what yields the best results. For both data sets, we first ran them through our quality control function. Since the filtered data already has all the empty droplets removed, this step will be skipped for the filtered data, but not for the raw data. In the following section we will quickly go over what steps we took in our quality control function and how we made sure they were sensible.

The first step in our pipeline is to seperate the antibody-derived tag (ADT) data from the remaining mRNA counts and clean them up. Ideally we'd like to identify the cells which failed to capture the ADT's and remove them. This is normally done by removing cells with ADT counts less than or equal to half of the total number of tags. If we do that we loose a huge number of cells as shown in the summary output. We also created a plot showing the distribution of the number of detected ADT's and from there we observe that we have lots of cells with zero ADT counts and a smaller peak at around 50. From this plot we decided to change the threshold at which cells will be discarded to 25, to not throw away as many cells. But even with this change it seems like to many cells will be discarded because of low quality ADT data, which is why we decided to ignore this part of the data altogether.

In the next step, the empty droplets will be removed. As mentioned before this only happens for the raw data. For this we ran the funciton emptyDrops(), which tests if the counts of a cell are significantly different from the ambient mRNA pool. If this is not the case, then it is most likely an empty droplet. After looking at some summary statistics of the output of this function, we realized that there were non-significant cellbarcodes, which were bounded by the number of iterations. We therefore increased the number of iterations in emptyDrops() until this wasn't the case anymore (100'000). There's also a plot showing the histogram of the p-values given by emptyDrops(), which should ideally be approximatelly uniform especially at the beginning. This was luckily the case. It is interesting to note, that this part of the quality control pipeline is responsible for the most amount of discarded cells.

Next we removed the dead cells, e.g. the cells with high mitochondrial gene expression. The approach we used was the same as in class. Lastly we also removed cells, which have low library sizes and detection rates. As seen in class we used the perCellQCMetrics() and isOutlier() functions to do so. Before removing all these cells we looked at the plots showing the above metrics to have an idea what they look like. We also produced a plot showing the log-fold-change of the discarded vs. kept cells. This plot should help decide if an entire cell type is accidently discarded because of our quality control. If that would be the case, we would observe an enrichment of a certain cell type in the discarded group, which can be identified by an increased expression of the corresponding marker genes.


```{r data-loading}
## data can be downloaded from : http://imlspenticton.uzh.ch/dump/files_for_levesque.tar

#TODO: can be deleted?
# patient1_HS.path <- file.path("data", "patient1_HS")
# 
# # read in filtered data
# fnameHS <- file.path(patient1_HS.path, "outs/filtered_feature_bc_matrix")


#read 10x data that has empty drops filtered but nothing else(?)
read_10x <- function() {
  sample_names <- list("patient1_HS", "patient1_SCC", "patient2_HS", "patient2_AK")
  patient1_HS.path <- file.path("data", "patient1_HS")
  patient1_SCC.path <- file.path("data", "patient1_SCC")
  patient2_HS.path <- file.path("data", "patient2_HS")
  patient2_AK.path <- file.path("data", "patient2_AK")
  
  
  # read in filtered data and create list of SingleCellExperiment objects
  paths <- list(patient1_HS.path, patient1_SCC.path, patient2_HS.path, patient2_AK.path)
  sces <- lapply(paths, function(i) read10xCounts(file.path(i, "outs/filtered_feature_bc_matrix"), col.names = TRUE))
  return(sces)
}
#read previusly stored data
read_previous_data <- function() {
  sces <- list()
  sces[[1]] <- readRDS('data/patient1_HS/clean_data/patient1_HS_cleanData_sce.RDS')
  sces[[2]] <- readRDS('data/patient1_SCC/clean_data/patient1_SCC_cleanData_sce.RDS')
  sces[[3]] <- readRDS('data/patient2_HS/clean_data/patient2_HS_cleanData_sce.RDS')
  sces[[4]] <- readRDS('data/patient2_AK/clean_data/patient2_AK_cleanData_sce.RDS')
  return(sces)
}
    
#read raw data and conduct QC
read_and_qc <- function(cores=4)
{
  sces <- list()
  sces[[1]] <- processCellRangerOutput("patient1_HS", 100000, TRUE, TRUE, TRUE, cores)
  sces[[2]] <- processCellRangerOutput("patient1_SCC", 100000, TRUE, TRUE, TRUE, cores)
  sces[[3]] <- processCellRangerOutput("patient2_HS", 100000, TRUE, TRUE, TRUE, cores)
  sces[[4]] <- processCellRangerOutput("patient2_AK", 100000, TRUE, TRUE, TRUE, cores)
  return(sces)
}

#pick your poison

sces <- read_previous_data()
#sces <- read_and_qc(6)
#sces <- read_10x()

```

```{r split-data}

sces <- lapply(sces, split_data)

# only select the gene expression data, drop the antibody capture
sce.patient1_HS <- sces[[1]]
sce.patient1_SCC <- sces[[2]]
sce.patient2_HS <- sces[[3]]
sce.patient2_AK <- sces[[4]]

# add name to sce for output graph
attr(sce.patient1_HS, "name") <- "Patient1 HS"
attr(sce.patient1_SCC, "name") <- "Patient1 SCC"
attr(sce.patient2_HS, "name") <- "Patient2 HS"
attr(sce.patient2_AK, "name") <- "Patient2 AK"

#convert rownames
# rownames(sce.patient1_HS) <- convert_rownames(sce.patient1_HS)
# rownames(sce.patient1_SCC) <- convert_rownames(sce.patient1_SCC)
# rownames(sce.patient2_HS) <- convert_rownames(sce.patient2_HS)
# rownames(sce.patient2_AK) <- convert_rownames(sce.patient2_AK)
```

## Normalization & reduction

```{r vst, include=FALSE}

results <- variance_stabilization(sce.patient1_HS, 2000)
sce.patient1_HS <- results[[1]]
hvg.patient1_HS <- results[[2]]

results <- variance_stabilization(sce.patient1_SCC, 2000)
sce.patient1_SCC <- results[[1]]
hvg.patient1_SCC <- results[[2]]

results <- variance_stabilization(sce.patient2_HS, 2000)
sce.patient2_HS <- results[[1]]
hvg.patient2_HS <- results[[2]]

results <- variance_stabilization(sce.patient2_AK, 2000)
sce.patient2_AK <- results[[1]]
hvg.patient2_AK <- results[[2]]
```

The markers found in the literature @scRNAseq @fibroblast were used to separate the different cell types and their subtypes within them. A few markers didn't appear at all in our data (WISP2 [Secretory-reticular fibroblast], SEPP1 [Macrophage], TYP1 [Melanocyte]) so the corresponding cell types were selected using their remaining markers present. The Pericyte and vSMC markers were removed because they clustered badly and polluted the Fibroblast population. Since those markers are known to correlate with specific cell types, their genes were added to the list of high variable genes used to produce a lower dimensional representation of the data, even when their expression didn't vary enough to be selected as highly variable.

```{r gene_table, echo=FALSE}
library(knitr)

cell_types <- c("Keratinocyte", "", "", "", "", "", "", "", "Fibroblast", "", "", "", "", "Pericyte & vSMC", "", "", "Endothelial cell", "", "Myeloid", "", "", "", "Lymphocyte", "Melanocyte", "Schwann cell", "Mitotic cell")

cell_subtypes <- c("", "basal", "suprabasal", "differentiated", "ORS", "channel", "sebaceous gland", "sweat gland", "", "secretory reticular", "proinflammatory", "secretory papillary", "mesenchymal", "", "pericyte", "vSMC", "", "lymphatic", "", "dendritic", "langerhans", "macrophage", "", "", "", "")

markers <- c(paste("DSC3","DSP","LGALS7"), paste("KRT5","KRT14","COL17A1"), paste("KRT1","KRT10"), paste("LOR", "SPINK5"), paste("KRT6B"), paste("GJB2","GJB6","ATP1B1"), paste("MGST1","FASN"), paste("DCD","AQP5"), paste("DCN", "LUM","MMP2"), paste("SLPI","CTHRC1","MFAP5","TSPAN8"), paste("CCL19","APOE","CXCL2","CXCL3","EFEMP1"), paste("APCDD1","ID1","WIF1","COL18A1","PTGDS"), paste("ASPN","POSTN","GPC3","TNN","SFRP1"), paste("ACTA2","TAGLN"), paste("RGS5"), paste("MYL9","TPM2","RERGL"), paste("PECAM1","SELE","CLDN5","VWF"), paste("PROX1","LYVE1"), paste("HLA-DRA","FCER1G","TYROBP","AIF1"), paste("CD1C"), paste("CD207"), paste("CD68","RNASE1","ITGAX"), paste("CD3D","CD3E","CD52","IL7R"), paste("DCT","MLANA","PMEL"), paste("CDH19","NGFR"), paste("UBE2C","PCNA"))

cell.df <- data.frame(Types=cell_types, Subtypes=cell_subtypes, Markers=markers)

kable(cell.df)
```

```{r pca}
# run PCA

# using known genes
# genes not found in the data : WISP2" "SEPP1" "TYP1" 
# genes linked to pericytevSMC : "ACTA2","TAGLN"
# were removed from the following list
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","DCN","LUM","MMP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","CDH19","NGFR","UBE2C","PCNA")

sce.patient1_HS <- sc_PCA(sce.patient1_HS, hvg.patient1_HS, known_markers)
sce.patient1_SCC <- sc_PCA(sce.patient1_SCC, hvg.patient1_SCC, known_markers)
sce.patient2_HS <- sc_PCA(sce.patient2_HS, hvg.patient2_HS, known_markers)
sce.patient2_AK <- sc_PCA(sce.patient2_AK, hvg.patient2_AK, known_markers)

```

## Clustering

```{r clustering}

results <- sc_cluster(sce.patient1_HS)
sce.patient1_HS <- results[[1]]
g.patient1_HS <- results[[2]]

results <- sc_cluster(sce.patient1_SCC)
sce.patient1_SCC <- results[[1]]
g.patient1_SCC <- results[[2]]

results <- sc_cluster(sce.patient2_HS)
sce.patient2_HS <- results[[1]]
g.patient2_HS <- results[[2]]

results <- sc_cluster(sce.patient2_AK)
sce.patient2_AK <- results[[1]]
g.patient2_AK <- results[[2]]

```

# Cluster annotation

## Known markers

```{r}
# genes not found in our data : "DCDN" "WISP2" "SEPP1" "TYP1" 
# were removed from the list below
genes <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
  # KERATINOCYTE
  keratinocyte = c("DSC3","DSP","LGALS7"),
  #keratinocyte_basal = c("KRT5","KRT14","COL17A1"),
  #keratinocyte_suprabasal = c("KRT1","KRT10"),
  #keratinocyte_differentiated = c("LOR", "SPINK5"),
  #keratinocyte_ORS = c("KRT6B"),
  #keratinocyte_channel = c("GJB2","GJB6","ATP1B1"),
  #keratinocyte_sebaceous_gland = c("MGST1","FASN"),
  #keratinocyte_sweat_gland  = c("DCD","AQP5"),
  # FIBROBLAST
  fibroblast  = c("DCN","LUM","MMP2"),
  # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
  #fibroblast_secretory_reticular = c(SLPI","CTHRC1","MFAP5","TSPAN8"),
  #fibroblast_proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
  #fibroblast_secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
  #fibroblast_mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1"),
  # PERICYTE & vSMC
  pericytevSMC  = c("ACTA2","TAGLN"), #vSMC : vascular smooth muscel cell
  #pericytevSMC_pericyte  = c("RGS5"),
  #pericytevSMC_vSMC  = c("MYL9","TPM2","RERGL"),
  # ENDOTHELIAL CELL
  endothelial  = c("PECAM1","SELE","CLDN5","VWF"),
  #endothelial_lymphatic  = c("PROX1","LYVE1"),
  # MYELOID CELL
  myeloid  = c("HLA-DRA","FCER1G","TYROBP","AIF1"),
  #myeloid_dendritic  = c("CD1C"),
  #myeloid_langerhans  = c("CD207"),
  #myeloid_macrophage = c("CD68","RNASE1","ITGAX"),
  # LYMPHOCYTE
  lymphocyte  = c("CD3D","CD3E","CD52","IL7R"),
  # MELANOCYTE
  melanocyte  = c("DCT","MLANA","PMEL"),
  # SCHWANN CELL
  schwann  = c("CDH19","NGFR"),
  # MITOTIC CELL
  mitotic  = c("UBE2C","PCNA")
)

# TODO: try to move the convert_rownames step higher in a block without output so we're less
# likely to run it multiple time (which would lead to weird output)
#rownames(sce.patient1_HS) <- convert_rownames(sce.patient1_HS, g.patient1_HS, genes)
results <- annotate_cells(sce.patient1_HS, g.patient1_HS, genes)
sce.patient1_HS <- results[[1]]
km.patient1_HS <- results[[2]]

#rownames(sce.patient1_SCC) <- convert_rownames(sce.patient1_SCC, g.patient1_SCC, genes)
results <- annotate_cells(sce.patient1_SCC, g.patient1_SCC, genes)
sce.patient1_SCC <- results[[1]]
km.patient1_SCC <- results[[2]]

#rownames(sce.patient2_HS) <- convert_rownames(sce.patient2_HS, g.patient2_HS, genes)
results <- annotate_cells(sce.patient2_HS, g.patient2_HS, genes)
sce.patient2_HS <- results[[1]]
km.patient2_HS <- results[[2]]

#rownames(sce.patient2_AK) <- convert_rownames(sce.patient2_AK, g.patient2_AK, genes)
results <- annotate_cells(sce.patient2_AK, g.patient2_AK, genes)
sce.patient2_AK <- results[[1]]
km.patient2_AK <- results[[2]]

```

## Pseudo-bulk aggregation

```{r}
#Patient 1 HS
#km.patient1_HS
#km.patient1_SCC
#rowData(sce.patient1_HS[c("keratinocyte", "fibroblast", "endothelial", "myeloid", )])$Symbol

results <- pseudobulk(sce.patient1_HS, km.patient1_HS)

sce.patient1_HS <- results[[1]]
pb.patient1_HS <- results[[2]]
mat.patient1_HS <- results[[3]]
cl2.patient1_HS <- results[[4]]

#Patient 1 SCC
results <- pseudobulk(sce.patient1_SCC, km.patient1_SCC)

sce.patient1_SCC <- results[[1]]
pb.patient1_SCC <- results[[2]]
mat.patient1_SCC <- results[[3]]
cl2.patient1_SCC <- results[[4]]

#Patient 2 HS
results <- pseudobulk(sce.patient2_HS, km.patient2_HS)

sce.patient2_HS <- results[[1]]
pb.patient2_HS <- results[[2]]
mat.patient2_HS <- results[[3]]
cl2.patient2_HS <- results[[4]]

#Patient 2 AK
results <- pseudobulk(sce.patient2_AK, km.patient2_AK)

sce.patient2_AK <- results[[1]]
pb.patient2_AK <- results[[2]]
mat.patient2_AK <- results[[3]]
cl2.patient2_AK <- results[[4]]

# create barplots of cell type proportions
df.celltypes.patient1 <- rbind(df_barplot_celltypes(sce.patient1_HS), df_barplot_celltypes(sce.patient1_SCC))
df.celltypes.patient1$Type <- c("HS", "HS", "HS", "HS", "HS", "HS", "HS", "HS", "SCC", "SCC", "SCC", "SCC", "SCC", "SCC", "SCC", "SCC")

df.celltypes.patient2 <- rbind(df_barplot_celltypes(sce.patient2_HS), df_barplot_celltypes(sce.patient2_AK))
df.celltypes.patient2$Type <- c("HS", "HS", "HS", "HS", "HS", "HS", "HS", "HS", "AK", "AK", "AK", "AK", "AK", "AK", "AK", "AK")
  
dynamic_barplot(df.celltypes.patient1, "Patient 1", T, "Proportion of cell types in", c(HS ='#999999', SCC = '#E69F00'))
dynamic_barplot(df.celltypes.patient2, "Patient 2", T, "Proportion of cell types in", c(HS ='#999999', AK = '#E69F00')))

```

# Sub population
```{r keratinocyte}

known_markers.kera <- c("KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5")

#annotate the subclusters
  genes.kera <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
    # KERATINOCYTE
    #keratinocyte = c("DSC3","DSP","LGALS7"),
    basal = c("KRT5","KRT14","COL17A1"),
    suprabasal = c("KRT1","KRT10"),
    differentiated = c("LOR", "SPINK5"),
    ORS = c("KRT6B"),
    channel = c("GJB2","GJB6","ATP1B1"),
    sebaceous_gland = c("MGST1","FASN"),
    sweat_gland  = c("DCD","AQP5")
    )
  
  
# TODO : find a way to avoid giving both known_merkers.kera and genes.kera since both contain the 
  # same information in essence. Maybe will be fixed by the first todo
  # yeah but then we are running an additional for loop/maybe we want different ones (hypothetically at least)
sce.patient1_HS.kera <- cluster_subtype(sce.patient1_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient1_SCC.kera <- cluster_subtype(sce.patient1_SCC, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_HS.kera <- cluster_subtype(sce.patient2_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_AK.kera <- cluster_subtype(sce.patient2_AK, "keratinocyte", known_markers.kera, genes.kera)

```

```{r kera_dyn}
genes.dyn <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0092867420306723
  keratinocyte_basal = "COL17A1",
  keratinocyte_cycling = "MKI67",
  keratinocyte_differentiating = "KRT1"
  )


# dynamic_plot(sce.patient1_HS.kera, g.patient1_HS.kera, genes.dyn)
# dynamic_plot(sce.patient1_SCC.kera, g.patient1_SCC.kera, genes.dyn)
# dynamic_plot(sce.patient2_HS.kera, g.patient2_HS.kera, genes.dyn)
# dynamic_plot(sce.patient2_AK.kera, g.patient2_AK.kera, genes.dyn)

## lelo_new plot

kera.dyn.patient1_HS <- dynamic_plot(sce.patient1_HS.kera, g.patient1_HS.kera, genes.dyn)
kera.dyn.patient1_SCC <- dynamic_plot(sce.patient1_SCC.kera, g.patient1_SCC.kera, genes.dyn)
kera.dyn.patient2_HS <- dynamic_plot(sce.patient2_HS.kera, g.patient2_HS.kera, genes.dyn)
kera.dyn.patient2_AK <- dynamic_plot(sce.patient2_AK.kera, g.patient2_AK.kera, genes.dyn)

kera.dyn.patient1 <- rbind(kera.dyn.patient1_HS, kera.dyn.patient1_SCC)
kera.dyn.patient1$Type <- c("HS", "HS", "HS", "SCC", "SCC", "SCC")

kera.dyn.patient2 <- rbind(kera.dyn.patient2_HS, kera.dyn.patient2_AK)
kera.dyn.patient2$Type <- c("HS", "HS", "HS", "AK", "AK", "AK")


dynamic_barplot(kera.dyn.patient1, "Patient 1", F, "Proportion of Keratinocyte states in", c(HS ='#999999', SCC = '#E69F00')))
dynamic_barplot(kera.dyn.patient2, "Patient 2", F, "Proportion of Keratinocyte states in", c(HS ='#999999', AK = '#E69F00')))

```

```{r fibroblast}

known_markers.fibro <- c("SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1")

genes.fibro <- list(
   #fibroblast  = c("DCN","LUM","MMP2"),
#   # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
   fibroblast_secretory_reticular = c("SLPI","CTHRC1","MFAP5","TSPAN8"),
   fibroblast_proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
   fibroblast_secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
   fibroblast_mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1")
)

sce.patient1_HS.fibro <- cluster_subtype(sce.patient1_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient1_SCC.fibro <- cluster_subtype(sce.patient1_SCC, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_HS.fibro <- cluster_subtype(sce.patient2_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_AK.fibro <- cluster_subtype(sce.patient2_AK, "fibroblast", known_markers.fibro, genes.fibro)

```


#######################################################################################
```{r check_markers, eval=FALSE}
# The following code was used to check which genes were missing from in the data
# the known_markers are from the litterature

data_markers <- rowData(sce.patient1_HS)$Symbol
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","DCN","LUM","MMP2","WISP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","ACTA2","TAGLN","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","SEPP1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","TYP1","CDH19","NGFR","UBE2C","PCNA")


idx_notfound <- which(!(known_markers %in% rowData(sce.patient1_HS)$Symbol))

print(paste("Done:", length(known_markers)-length(idx_notfound), "/", length(known_markers), "markers found"))
cat("Genes missing in dataset: ", known_markers[idx_notfound])
```
