---
title: "scRNAseq analysis of Levesque’s data"
author: "Leonor Schubert, Jonathan Haab, Flavio Rump"
date: "22 12 2020"
output: html_document
---
from pipeline example described by Pierre-Luc Germain, course sta426 UZH

# Loading necessary libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scran)
  library(scater)
  library(batchelor)
  library(scDblFinder)
  library(sctransform)
  library(muscat)
  library(SEtools)
  library(cowplot)
  library(BiocParallel)
  library(ComplexHeatmap)
  
  library(DropletUtils) # for read10xCounts
  library(Matrix) # to handle sparse matrix
  library(BiocNeighbors) # for kNN graph
  library(igraph) # for cluster_louvain
  
  library(easyGgplot2)
})
```

# Preprocessing & Clustering

## Loading the data

```{r data-loading}
## data can be downloaded from : http://imlspenticton.uzh.ch/dump/files_for_levesque.tar

# set local path
local.path <- getwd()
setwd(local.path)

#define data we want to look at
sample_names <- list("patient1_HS", "patient1_SCC", "patient2_HS", "patient2_AK")
patient1_HS.path <- file.path("data", "patient1_HS")
patient1_SCC.path <- file.path("data", "patient1_SCC")
patient2_HS.path <- file.path("data", "patient2_HS")
patient2_AK.path <- file.path("data", "patient2_AK")

#read 10X data

# read in filtered data and create list of SingleCellExperiment objects
paths <- list(patient1_HS.path, patient1_SCC.path, patient2_HS.path, patient2_AK.path)
#sces <- lapply(paths, function(i) read10xCounts(file.path(i, "outs/filtered_feature_bc_matrix"), col.names = TRUE))

#read our own data
sces <- list()
sces[[1]] <- readRDS('clean_data/patient1_HS_cleanData_sce.RDS')
sces[[2]] <- readRDS('clean_data/patient1_SCC_cleanData_sce.RDS')
sces[[3]] <- readRDS('clean_data/patient2_HS_cleanData_sce.RDS')
sces[[4]] <- readRDS('clean_data/patient2_AK_cleanData_sce.RDS')

#read 10X data
# read in filtered data and create list of SingleCellExperiment objects
paths <- list(patient1_HS.path, patient1_SCC.path, patient2_HS.path, patient2_AK.path)
#sces <- lapply(paths, function(i) read10xCounts(file.path(i, "outs/filtered_feature_bc_matrix"), col.names = TRUE))

#read our own data
sces <- list()
sces[[1]] <- readRDS('clean_data/patient1_HS_cleanData_sce.RDS')
sces[[2]] <- readRDS('clean_data/patient1_SCC_cleanData_sce.RDS')
sces[[3]] <- readRDS('clean_data/patient2_HS_cleanData_sce.RDS')
sces[[4]] <- readRDS('clean_data/patient2_AK_cleanData_sce.RDS')

```

```{r split-data}

split_data <- function(sceo) {
  # Split the data, store ADT in alternative experiment
  sceo <- splitAltExps(sceo, rowData(sceo)$Type)
  
  # Coerce sparse matrix for ADT into a dense matrix
  counts(altExp(sceo)) <- as.matrix(counts(altExp(sceo)))
  sceo
}

sces <- lapply(sces, split_data)

# only select the gene expression data, drop the antibody capture
sce.patient1_HS <- sces[[1]]
sce.patient1_SCC <- sces[[2]]
sce.patient2_HS <- sces[[3]]
sce.patient2_AK <- sces[[4]]

# add name to sce for output graph
attr(sce.patient1_HS, "name") <- "Patient1 HS"
attr(sce.patient1_SCC, "name") <- "Patient1 SCC"
attr(sce.patient2_HS, "name") <- "Patient2 HS"
attr(sce.patient2_AK, "name") <- "Patient2 AK"
```


## Normalization & reduction

```{r vst, include=FALSE}
# get rid of seldom detected genes
remove_rare_genes <- function(sceo, num_genes) {
  sceo <- sceo[(rowSums(counts(sceo) > 0) > num_genes),]
}

variance_stabilization <- function(sceo, num_genes) {

  sceo <- remove_rare_genes(sceo, 4)
  vsto <- suppressWarnings(sctransform::vst(counts(sceo)))
  logcounts(sceo, withDimnames=FALSE) <- vsto$y
  
  # Check that new assay was added to sce
  #assays(sce.patient1_HS)
  
  # get highly-variable genes
  hvgo <- row.names(sceo)[order(vsto$gene_attr$residual_variance, decreasing=TRUE)[1:num_genes]]
  results <- list("sceo" <- sceo, "hvgo" <- hvgo)
  return(results)
}

results <- variance_stabilization(sce.patient1_HS, 2000)
sce.patient1_HS <- results[[1]]
hvg.patient1_HS <- results[[2]]

results <- variance_stabilization(sce.patient1_SCC, 2000)
sce.patient1_SCC <- results[[1]]
hvg.patient1_SCC <- results[[2]]

results <- variance_stabilization(sce.patient2_HS, 2000)
sce.patient2_HS <- results[[1]]
hvg.patient2_HS <- results[[2]]

results <- variance_stabilization(sce.patient2_AK, 2000)
sce.patient2_AK <- results[[1]]
hvg.patient2_AK <- results[[2]]
```


ADD A NICE TABLE WITH ALL THE MARKER USED (and their source) AND EXPLAINED WHICH WEREN'T FOUND
also explain that we decided to add the known markers from the litterature into the group of genes used to produce a low-dimensional representation of the data even when they were not part of the most variable genes

```{r pca}
# run PCA

# using known genes
# genes not found in the data : "DCDN" "WISP2" "SEPP1" "TYP1" 
# genes linked to pericytevSMC : "ACTA2","TAGLN"
# were removed from the following list
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","LUM","MMP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","CDH19","NGFR","UBE2C","PCNA")


sc_PCA <- function(sceo, hvgo, known_markers) {
  # only gives index of the first encountered
  known_genes <- rownames(sceo)[which(rowData(sceo)$Symbol %in% known_markers)]
  
  # Check that the genes we know are also part of the highly variable genes
  idx_notfound <- which(!(known_genes %in% hvgo))
  
  # print the markers that were not found in the data
  #known_markers[idx_notfound]
  
  # if some of the known markers were not selected, add them
  if (length(idx_notfound) > 0) {
    cat("Adding", known_markers[idx_notfound], "to the gene set used for PCA in", attr(sceo, "name"), "\n")
    hvgo <- c(hvgo, known_genes[idx_notfound])
  }
    
  # using highly variable genes
  sceo <- runPCA(sceo, subset_row=hvgo)
  
  # check the variance explained by the PCs:
  pc.var <- attr(reducedDim(sceo),"percentVar")
  plot(pc.var, xlab="PCs", ylab="% variance explained", main=paste("Variance explained across PCs for", attr(sceo, "name")))
  
  # restrict to the first 20 components:
  reducedDim(sceo) <- reducedDim(sceo)[,1:20]
  
  # run TSNE and UMAP based on the PCA:
  sceo <- runTSNE(sceo, dimred="PCA")
  sceo <- runUMAP(sceo, dimred="PCA")
}

sce.patient1_HS <- sc_PCA(sce.patient1_HS, hvg.patient1_HS, known_markers)
sce.patient1_SCC <- sc_PCA(sce.patient1_SCC, hvg.patient1_SCC, known_markers)
sce.patient2_HS <- sc_PCA(sce.patient2_HS, hvg.patient2_HS, known_markers)
sce.patient2_AK <- sc_PCA(sce.patient2_AK, hvg.patient2_AK, known_markers)

```

## Clustering

```{r clustering}
sc_cluster <- function(sceo, k=30) {
  go <- buildKNNGraph(sceo, BNPARAM=AnnoyParam(), use.dimred="PCA", k=k)

  sceo$cluster <- as.factor(cluster_louvain(go)$membership)

  #plots <- plot_grid( plotTSNE(sceo, colour_by="cluster", text_by="cluster") + ggtitle(attr(sceo, "name")), plotUMAP(sceo, colour_by="cluster", text_by="cluster") + ggtitle(attr(sceo, "name")) ) 
  
  plots <- plot_grid( 
    plotTSNE(sceo, colour_by="cluster", text_by="cluster"), 
    plotUMAP(sceo, colour_by="cluster", text_by="cluster"))
  title <- ggdraw() + draw_label(attr(sceo, "name"), fontface='bold')
  plots <- plot_grid(title, plots, ncol=1, rel_heights=c(0.1, 1))
  print(plots)
  
  return(list(sceo, go))
}

results <- sc_cluster(sce.patient1_HS)
sce.patient1_HS <- results[[1]]
g.patient1_HS <- results[[2]]

results <- sc_cluster(sce.patient1_SCC)
sce.patient1_SCC <- results[[1]]
g.patient1_SCC <- results[[2]]

results <- sc_cluster(sce.patient2_HS)
sce.patient2_HS <- results[[1]]
g.patient2_HS <- results[[2]]

results <- sc_cluster(sce.patient2_AK)
sce.patient2_AK <- results[[1]]
g.patient2_AK <- results[[2]]

```

# Cluster annotation

## Known markers

```{r}
# genes not found in our data : "DCDN" "WISP2" "SEPP1" "TYP1" 
# were removed from the list below
genes <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
  # KERATINOCYTE
  keratinocyte = c("DSC3","DSP","LGALS7"),
  #keratinocyte_basal = c("KRT5","KRT14","COL17A1"),
  #keratinocyte_suprabasal = c("KRT1","KRT10"),
  #keratinocyte_differentiated = c("LOR", "SPINK5"),
  #keratinocyte_ORS = c("KRT6B"),
  #keratinocyte_channel = c("GJB2","GJB6","ATP1B1"),
  #keratinocyte_sebaceous_gland = c("MGST1","FASN"),
  #keratinocyte_sweat_gland  = c("DCD","AQP5"),
  # FIBROBLAST
  fibroblast  = c("LUM","MMP2"),
  # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
  #fibroblast_secretory_reticular = c(SLPI","CTHRC1","MFAP5","TSPAN8"),
  #fibroblast_proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
  #fibroblast_secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
  #fibroblast_mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1"),
  # PERICYTE & vSMC
  #pericytevSMC  = c("ACTA2","TAGLN"), #vSMC : vascular smooth muscel cell
  #pericytevSMC_pericyte  = c("RGS5"),
  #pericytevSMC_vSMC  = c("MYL9","TPM2","RERGL"),
  # ENDOTHELIAL CELL
  endothelial  = c("PECAM1","SELE","CLDN5","VWF"),
  #endothelial_lymphatic  = c("PROX1","LYVE1"),
  # MYELOID CELL
  myeloid  = c("HLA-DRA","FCER1G","TYROBP","AIF1"),
  #myeloid_dendritic  = c("CD1C"),
  #myeloid_langerhans  = c("CD207"),
  #myeloid_macrophage = c("CD68","RNASE1","ITGAX"),
  # LYMPHOCYTE
  lymphocyte  = c("CD3D","CD3E","CD52","IL7R"),
  # MELANOCYTE
  melanocyte  = c("DCT","MLANA","PMEL"),
  # SCHWANN CELL
  schwann  = c("CDH19","NGFR"),
  # MITOTIC CELL
  mitotic  = c("UBE2C","PCNA")
)

convert_rownames <- function(sceo, go, genes) {
  return(paste(rownames(sceo), rowData(sceo)$Symbol, sep = "."))
}

annotate_cells <- function(sceo, go, genes) {
  kmo <- lapply(genes, FUN=function(go) grep(paste0(go, "$", collapse="|"), rownames(sceo), value=TRUE))
  
  # TODO: try to modify this so that we don't have to convert the rownames anymore and use
  # the marker names save in rowData(sceo)$Symbol directly
  #kmo <- lapply(genes, FUN=function(go) grep(paste0(go, "$", collapse="|"), rowData(sceo)$Symbol, value=TRUE))
  #print(kmo)
  
  return(list(sceo, kmo))
}


rownames(sce.patient1_HS) <- convert_rownames(sce.patient1_HS, g.patient1_HS, genes)
results <- annotate_cells(sce.patient1_HS, g.patient1_HS, genes)
sce.patient1_HS <- results[[1]]
km.patient1_HS <- results[[2]]

rownames(sce.patient1_SCC) <- convert_rownames(sce.patient1_SCC, g.patient1_SCC, genes)
results <- annotate_cells(sce.patient1_SCC, g.patient1_SCC, genes)
sce.patient1_SCC <- results[[1]]
km.patient1_SCC <- results[[2]]

rownames(sce.patient2_HS) <- convert_rownames(sce.patient2_HS, g.patient2_HS, genes)
results <- annotate_cells(sce.patient2_HS, g.patient2_HS, genes)
sce.patient2_HS <- results[[1]]
km.patient2_HS <- results[[2]]

rownames(sce.patient2_AK) <- convert_rownames(sce.patient2_AK, g.patient2_AK, genes)
results <- annotate_cells(sce.patient2_AK, g.patient2_AK, genes)
sce.patient2_AK <- results[[1]]
km.patient2_AK <- results[[2]]

```

## Pseudo-bulk aggregation

```{r}
# mean logcounts by cluster
pseudobulk <- function(sceo, kmo) {
  pbo <- aggregateData(sceo, "logcounts", by=c("cluster"), fun="mean")

  # TODO : find a way to delete the left annotation which is unreadable
  
  # build a heatmap of the mean logcounts of the known markers:
  h <- pheatmap(assay(pbo)[unlist(kmo),], annotation_row=data.frame(row.names=unlist(kmo), type=rep(names(kmo), lengths(kmo))), split=rep(names(kmo), lengths(kmo)), annotation_names_row=FALSE, cluster_rows=FALSE, scale="row", main=paste(attr(sceo, "name"),"before markers aggregation"), fontsize_row=6, fontsize_col=10, angle_col = "45")
  print(h)
  
  #--- aggregation markers
  # we will assign clusters to the cell type whose markers are the most expressed
  
  # we extract the pseudo-bulk counts of the markers for each cluster
  mato <- assay(pbo)[unlist(kmo),]
  
  # we aggregate across markers of the same type
  mato <- aggregate(t(scale(t(mato))), by=list(type=rep(names(kmo), lengths(kmo))), FUN=sum)
  
  # for each column (cluster), we select the row (cell type) which has the maximum aggregated value
  cl2o<- mato[,1][apply(mato[,-1], 2, FUN=which.max)]
  # we convert the cells' cluster labels to cell type labels:
  sceo$cluster2 <- cl2o[sceo$cluster]
  
  # we aggregate again to pseudo-bulk using the new clusters
  pbo <- aggregateData(sceo, "logcounts", by=c("cluster2"), fun="mean")
  
  # we plot again the expression of the markers as a sanity check
  
  # If we want to hide the gene markers show_rownames = FALSE
  h1 <- pheatmap(assay(pbo)[unlist(kmo),], annotation_row=data.frame(row.names=unlist(kmo), type=rep(names(kmo), lengths(kmo))), annotation_names_row=FALSE, split=rep(names(kmo), lengths(kmo)), cluster_rows=FALSE, scale="row", main=paste(attr(sceo, "name"),"after markers aggregation"), fontsize_row=6, fontsize_col=10, angle_col = "45")
  print(h1)
  
  # UMAP plot
  p <- plotUMAP(sceo, colour_by="cluster2", text_by="cluster2", point_size=1) + ggtitle(attr(sceo, "name")) + theme(legend.title=element_blank())
  print(p)
  
  return(list(sceo, pbo, mato, cl2o))
}

#Patient 1 HS
#km.patient1_HS
#km.patient1_SCC
#rowData(sce.patient1_HS[c("keratinocyte", "fibroblast", "endothelial", "myeloid", )])$Symbol

results <- pseudobulk(sce.patient1_HS, km.patient1_HS)

sce.patient1_HS <- results[[1]]
pb.patient1_HS <- results[[2]]
mat.patient1_HS <- results[[3]]
cl2.patient1_HS <- results[[4]]

#Patient 1 SCC
results <- pseudobulk(sce.patient1_SCC, km.patient1_SCC)

sce.patient1_SCC <- results[[1]]
pb.patient1_SCC <- results[[2]]
mat.patient1_SCC <- results[[3]]
cl2.patient1_SCC <- results[[4]]

#Patient 2 HS
results <- pseudobulk(sce.patient2_HS, km.patient2_HS)

sce.patient2_HS <- results[[1]]
pb.patient2_HS <- results[[2]]
mat.patient2_HS <- results[[3]]
cl2.patient2_HS <- results[[4]]

#Patient 2 AK
results <- pseudobulk(sce.patient2_AK, km.patient2_AK)

sce.patient2_AK <- results[[1]]
pb.patient2_AK <- results[[2]]
mat.patient2_AK <- results[[3]]
cl2.patient2_AK <- results[[4]]

```

# Sub population
```{r keratinocyte}


cluster_subtype <- function(sceo, cell_type, known_markers, subtype_markers) {
  # select the cell labeled as keratinocytes in the previous step
  sceo.sub <- sceo[,sceo$cluster2==cell_type]
  
  sceo.sub <- remove_rare_genes(sceo.sub, 4)
  results <- variance_stabilization(sceo.sub, 2000)
  
  sceo.sub <- results[[1]]
  hvgo.sub <- results[[2]]
  
  #---- run the pipeline again on that subdataset
  sceo.sub <- sc_PCA(sceo.sub, hvgo.sub, known_markers)
  
  #--- clustering
  
  results <- sc_cluster(sceo.sub)
  sceo.sub <- results[[1]]
  go.sub <- results[[2]]
  
  results <- annotate_cells(sceo.sub, go.sub, subtype_markers)
  
  sceo.sub <- results[[1]]
  kmo.sub <- results[[2]]
  
  #---
  
  results <- pseudobulk(sceo.sub, kmo.sub)
  
  sceo.sub <- results[[1]]
  pbo.sub <- results[[2]]
  mato.sub <- results[[3]]
  cl2o.sub <- results[[4]]
  
  return(sceo.sub)

}


known_markers.kera <- c("KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5")

#annotate the subclusters
  genes.kera <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0923181120301985?via%3Dihub#bib0050
    # KERATINOCYTE
    #keratinocyte = c("DSC3","DSP","LGALS7"),
    basal = c("KRT5","KRT14","COL17A1"),
    suprabasal = c("KRT1","KRT10"),
    differentiated = c("LOR", "SPINK5"),
    ORS = c("KRT6B"),
    channel = c("GJB2","GJB6","ATP1B1"),
    sebaceous_gland = c("MGST1","FASN"),
    sweat_gland  = c("DCD","AQP5")
    )
  
  
# TODO : find a way to avoid giving both known_merkers.kera and genes.kera since both contain the 
  # same information in essence. Maybe will be fixed by the first todo
sce.patient1_HS.kera <- cluster_subtype(sce.patient1_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient1_SCC.kera <- cluster_subtype(sce.patient1_SCC, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_HS.kera <- cluster_subtype(sce.patient2_HS, "keratinocyte", known_markers.kera, genes.kera)
sce.patient2_AK.kera <- cluster_subtype(sce.patient2_AK, "keratinocyte", known_markers.kera, genes.kera)

```

```{r kera_dyn}
# need the previous block to be runned (at least until l. 298)

genes.dyn <- list(
  # classification: https://www.sciencedirect.com/science/article/pii/S0092867420306723
  keratinocyte_basal = "COL17A1",
  keratinocyte_cycling = "MKI67",
  keratinocyte_differentiating = "KRT1"
  )

dynamic_plot <- function(sceo, go, genes) {
    kmo <- lapply(genes, FUN=function(go) grep(paste0(go, "$", collapse="|"), rownames(sceo), value=TRUE))
  
  # mean logcounts by cluster:
  pbo <- aggregateData(sceo, "logcounts", by=c("cluster"), fun="mean")
  lengths(kmo)
  h <- pheatmap(assay(pbo)[unlist(kmo),], annotation_row=data.frame(row.names=unlist(kmo), type=rep(names(kmo), lengths(kmo))), split=rep(names(kmo), lengths(kmo)), cluster_rows=FALSE, scale="row", main="Before markers aggregation", fontsize_row=6, fontsize_col=10)
  h
  
  # we will assign clusters to the cell type whose markers are the most expressed
  # we extract the pseudo-bulk counts of the markers for each cluster
  mato <- assay(pbo)[unlist(kmo),]
  
  # we aggregate across markers of the same type
  mato <- aggregate(t(scale(t(mato))), by=list(type=rep(names(kmo), lengths(kmo))), FUN=sum)
  
  # for each column (cluster), we select the row (cell type) which has the maximum aggregated value
  cl3o <- mato[,1][apply(mato[,-1], 2, FUN=which.max)]
  
  # we convert the cells' cluster labels to cell type labels:
  sceo$cluster3 <- cl3o[sceo$cluster]
  
  print(sceo)
  
  # we aggregate again to pseudo-bulk using the new clusters
  pbo <- aggregateData(sceo, "logcounts", by=c("cluster3"), fun="mean")
  # we plot again the expression of the markers as a sanity check
  h1 <- pheatmap(assay(pbo)[unlist(kmo),], annotation_row=data.frame(row.names=unlist(kmo), type=rep(names(kmo), lengths(kmo))), split=rep(names(kmo), lengths(kmo)), cluster_rows=FALSE, scale="row", main="After markers aggregation", fontsize_row=6, fontsize_col=10) #, scale="row", main="After markers aggregation", fontsize_row=6, fontsize_col=10)
  h1
  plotUMAP(sceo, colour_by="cluster3", text_by="cluster3", point_size=1)
  
  
  #---- Histograms
  total <- ncol(sceo)
  basal <- ncol(sceo[,sceo$cluster3=="keratinocyte_basal"])
  cycling <- ncol(sceo[,sceo$cluster3=="keratinocyte_cycling"])
  diff <- ncol(sceo[,sceo$cluster3=="keratinocyte_differentiating"])
  counts <- c(basal, cycling, diff)
  percentage <- counts/total
  kera.dyn <- data.frame(names=c("Basal", "Cycling", "Differentiating"), percentage=percentage)
  kera.dyn
  
  #p <- ggplot(data=kera.dyn, aes(x=names, y=percentage, fill=names)) + geom_bar(stat="identity") + labs(title=paste("Proportion of Keratinocyte states in", attr(sceo, "name")),x="Subtypes", y="Percentage") + ylim(0,1) + theme(legend.position='none')
  #print(p)
  return(kera.dyn)
}

# dynamic_plot(sce.patient1_HS.kera, g.patient1_HS.kera, genes.dyn)
# dynamic_plot(sce.patient1_SCC.kera, g.patient1_SCC.kera, genes.dyn)
# dynamic_plot(sce.patient2_HS.kera, g.patient2_HS.kera, genes.dyn)
# dynamic_plot(sce.patient2_AK.kera, g.patient2_AK.kera, genes.dyn)

## lelo_new plot

kera.dyn.patient1_HS <- dynamic_plot(sce.patient1_HS.kera, g.patient1_HS.kera, genes.dyn)
kera.dyn.patient1_SCC <- dynamic_plot(sce.patient1_SCC.kera, g.patient1_SCC.kera, genes.dyn)
kera.dyn.patient2_HS <- dynamic_plot(sce.patient2_HS.kera, g.patient2_HS.kera, genes.dyn)
kera.dyn.patient2_AK <- dynamic_plot(sce.patient2_AK.kera, g.patient2_AK.kera, genes.dyn)

kera.dyn.patient1 <- rbind(kera.dyn.patient1_HS, kera.dyn.patient1_SCC)
kera.dyn.patient1$Type <- c("HS", "HS", "HS", "SCC", "SCC", "SCC")

kera.dyn.patient2 <- rbind(kera.dyn.patient2_HS, kera.dyn.patient2_AK)
kera.dyn.patient2$Type <- c("HS", "HS", "HS", "AK", "AK", "AK")

dynamic_barplot <- function(df, name){
  
  p <- ggplot2.barplot(data=df, xName="names", yName="percentage",
      groupName="type", 
      position=position_dodge(),
      #background and line colors
      backgroundColor="white", color="black", 
      xtitle="Subtypes", ytitle="Percentage", 
      mainTitle=paste("Proportion of Keratinocyte states in", name),
      removePanelGrid=TRUE, removePanelBorder=TRUE,
      axisLine=c(0.5, "solid", "black"),
      groupColors=c('#999999','#E69F00'),
      ) 
  print(p)
}

dynamic_barplot(kera.dyn.patient1, "Patient 1")
dynamic_barplot(kera.dyn.patient2, "Patient 2")

```

```{r fibroblast}

known_markers.fibro <- c("SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1")

genes.fibro <- list(
   #fibroblast  = c("LUM","MMP2"),
#   # fibroblast subclassification  from: https://www.nature.com/articles/s42003-020-0922-4#
   secretory_reticular = c("SLPI","CTHRC1","MFAP5","TSPAN8"),
   proinflammatory = c("CCL19","APOE","CXCL2","CXCL3","EFEMP1"),
   secretory_papillary = c("APCDD1","ID1","WIF1","COL18A1","PTGDS"),
   mesenchymal = c("ASPN","POSTN","GPC3","TNN","SFRP1")
)

sce.patient1_HS.fibro <- cluster_subtype(sce.patient1_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient1_SCC.fibro <- cluster_subtype(sce.patient1_SCC, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_HS.fibro <- cluster_subtype(sce.patient2_HS, "fibroblast", known_markers.fibro, genes.fibro)
sce.patient2_AK.fibro <- cluster_subtype(sce.patient2_AK, "fibroblast", known_markers.fibro, genes.fibro)

```

```{r check_markers, eval=FALSE}
# The following code was used to check which genes were missing from in the data
# the known_markers are from the litterature

data_markers <- rowData(sce.patient1_HS)$Symbol
known_markers <- c("DSC3","DSP","LGALS7","KRT5","KRT14","COL17A1","KRT1","KRT10","LOR", "SPINK5","KRT6B","GJB2","GJB6","ATP1B1","MGST1","FASN","DCD","AQP5","DCDN","LUM","MMP2","WISP2","SLPI","CTHRC1","MFAP5","TSPAN8","CCL19","APOE","CXCL2","CXCL3","EFEMP1","APCDD1","ID1","WIF1","COL18A1","PTGDS","ASPN","POSTN","GPC3","TNN","SFRP1","ACTA2","TAGLN","RGS5","MYL9","TPM2","RERGL","PECAM1","SELE","CLDN5","VWF","PROX1","LYVE1","HLA-DRA","FCER1G","TYROBP","AIF1","CD1C","CD207","CD68","RNASE1","SEPP1","ITGAX","CD3D","CD3E","CD52","IL7R","DCT","MLANA","PMEL","TYP1","CDH19","NGFR","UBE2C","PCNA")


idx_notfound <- which(!(known_markers %in% rowData(sce.patient1_HS)$Symbol))

print(paste("Done:", length(known_markers)-length(idx_notfound), "/", length(known_markers), "markers found"))
cat("Genes missing in dataset: ", known_markers[idx_notfound])
```
